---
title: "understanding data visualization"
author: "Ashish Dutt"
date: "February 8, 2018"
output: pdf_document
reference: http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#Jitter%20Plot
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to data visualization
With ever increasing volume of data, it is impossible to tell stories without visualizations. Data visualization is an art of how to turn numbers into useful knowledge. Before the technical implementations of the visualization, let’s see first how to select the right chart type.

An effective chart is one that:

1. Conveys the right information without distorting facts.
2 Is simple but elegant. It should not force you to think much in order to get it.
3. Aesthetics supports information rather that overshadow it.
4. Is not overloaded with information.

### Selecting the Right Chart Type
Primarily, there are 8 types of objectives you may construct plots. So, before you actually make the plot, try and figure what findings and relationships you would like to convey or examine through the visualization. Chances are it will fall under one (or sometimes more) of these 8 categories.

```{r}
knitr::include_graphics("C:\\Users\\ashish.dutt\\Downloads\\choosing-a-good-chart.pdf")

```

In your day-to-day activities, you’ll come across the below listed 7 charts most of the time.

1. Correlation
  + Scatterplot
  + Scatterplot With Encircling
  + Jitter Plot
  + Bubble Plot
  + Marginal Histogram / Boxplot
2. Deviation
  + Diverging Bars
  + Area Chart
3. Ranking
  + Ordered Bar Chart
  + Dot Plot
  + Slope Chart
4. Distribution
  + Histogram
  + Density Plot
  + Box Plot
  + Dot + Box Plot
5. Composition
  + Waffle Chart
  + Bar Chart
6. Change
  + Time Series Plots
  + Stacked Area Chart
  + Seasonal Plot
7. Groups
  + Dendrogram
  + Clusters
8. Spatial
+ Open Street Map
+ Google Road Map

Load the required libraries
```{r}
options(scipen = 999) # turn-off scientific notation like 1e+48
library(ggplot2)
library(ggalt)
library(gapminder)
library(ggExtra)
library(ggcorrplot)
library(quantmod)
library(xts)
library(scales)
library(ggthemes)
library(lubridate)
library(ggdendro)
library(ggmap)
```

### 1. Correlation

#### Scatter Plot

**When to use**: Scatter Plot is used to see the relationship between two continuous variables.

It can be drawn using `geom_point()`. Additionally, `geom_smooth` which draws a smoothing line (based on loess) by default, can be tweaked to draw the line of best fit by setting `method='lm'`.

```{r}
theme_set(theme_bw())  # pre-set the bw theme.
data("midwest", package = "ggplot2")
# midwest <- read.csv("http://goo.gl/G1K41K")  # bkup data source

# Scatterplot
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + 
  coord_cartesian(ylim = c(0, 500000))+ 
  labs(subtitle="Area Vs Population", 
       y="Population", 
       x="Area", 
       title="Scatterplot", 
       caption = "Source: midwest")

plot(gg)
```
A categorical variable can be used in a scatterplot to show the characterstics of the plotted variables. So, the above code can be modified as,

#### Scatterplot With Encircling
When presenting the results, sometimes I would encirlce certain special group of points or region in the chart so as to draw the attention to those peculiar cases. This can be conveniently done using the `geom_encircle()` in `ggalt` package.

Within  `geom_encircle()` , set the `data` to a new dataframe that contains only the points (rows) or interest. Moreover, You can `expand` the curve so as to pass just outside the points. The `color` and `size` (thickness) of the curve can be modified as well. See below example.

```{r}

midwest_select <- midwest[midwest$poptotal > 350000 & 
                            midwest$poptotal <= 500000 & 
                            midwest$area > 0.01 & 
                            midwest$area < 0.1, ]

# Plot
ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) +   # draw points
  geom_smooth(method="loess", se=F) + 
  #xlim(c(0, 0.6)) + 
  #ylim(c(0, 500000)) +   # draw smoothing line
  coord_cartesian(ylim = c(0, 500000))+
  geom_encircle(aes(x=area, y=poptotal), 
                data=midwest_select, 
                color="red", 
                size=2, 
                expand=0.06) +   # encircle
  labs(subtitle="Area Vs Population", 
       y="Population", 
       x="Area", 
       title="Scatterplot + Encircle", 
       caption="Source: midwest")

```
#### Jitter Plot
Let’s look at a new data to draw the scatterplot. This time, I will use the `mpg` dataset to plot city mileage (`cty`) vs highway mileage (`hwy`).

```{r}
g <- ggplot(data=mpg, aes(cty, hwy))

# Scatterplot
g + geom_jitter(width = 0.5, size=1) + 
  geom_smooth(method="lm", se=F) +
  labs(subtitle="mpg: city vs highway mileage", 
       y="hwy", 
       x="cty", 
       title="Scatterplot with overlapping points", 
       caption="Source: midwest")
```

#### Bubble Plot

While scatterplot lets you compare the relationship between 2 continuous variables, bubble chart serves well if you want to understand relationship within the underlying groups based on:

1. A Categorical variable (by changing the color) and
2. Another continuous variable (by changing the size of points).

In simpler words, bubble charts are more suitable if you have 4-Dimensional data where two of them are numeric (X and Y) and one other categorical (color) and another numeric variable (size).

```{r}
mpg_select <- mpg[mpg$manufacturer %in% c("audi", "ford", "honda", "hyundai"), ]

# Scatterplot
g <- ggplot(mpg_select, aes(displ, cty)) + 
  labs(subtitle="mpg: Displacement vs City Mileage",
       title="Bubble chart")

g + geom_jitter(aes(col=manufacturer, size=hwy)) + 
  geom_smooth(aes(col=manufacturer), method="lm", se=F)

```
#### Marginal Histogram / Boxplot

If you want to show the relationship as well as the distribution in the same chart, use the marginal histogram. It has a histogram of the X and Y variables at the margins of the scatterplot.

This can be implemented using the `ggMarginal()` function from the `‘ggExtra’` package. Apart from a `histogram`, you could choose to draw a `marginal boxplot` or `density` plot by setting the respective `type` option.

```{r}
g <- ggplot(mpg, aes(cty, hwy)) + 
  geom_count() + 
  geom_smooth(method="lm", se=F)

ggMarginal(g, type = "histogram", fill="transparent")
#plt1<- ggMarginal(g, type = "boxplot", fill="transparent")

```

### 2. Deviation
Compare variation in values between small number of items (or categories) with respect to a fixed reference.


#### Diverging Bars
Diverging Bars is a bar chart that can handle both negative and positive values. This can be implemented by a smart tweak with `geom_bar()`. But the usage of `geom_bar()` can be quite confusing. Thats because, it can be used to make a bar chart as well as a histogram. Let me explain.

By default, `geom_bar()` has the stat set to count. That means, when you provide just a continuous X variable (and no Y variable), it tries to make a histogram out of the data.

In order to make a bar chart create bars instead of histogram, you need to do two things.

a. Set `stat=identity`

b. Provide both `x` and `y` inside `aes()` where, x is either `character` or `factor` and y is numeric.
In order to make sure you get diverging bars instead of just bars, make sure, your categorical variable has 2 categories that changes values at a certain threshold of the continuous variable. In below example, the mpg from mtcars dataset is normalised by computing the z score. Those vehicles with mpg above zero are marked green and those below are marked red.

```{r}
# Data Prep
data("mtcars")  # load data
mtcars$`car name` <- rownames(mtcars)  # create new column for car names
mtcars$mpg_z <- round((mtcars$mpg - mean(mtcars$mpg))/sd(mtcars$mpg), 2)  # compute normalized mpg
mtcars$mpg_type <- ifelse(mtcars$mpg_z < 0, "below", "above")  # above / below avg flag
mtcars <- mtcars[order(mtcars$mpg_z), ]  # sort
mtcars$`car name` <- factor(mtcars$`car name`, levels = mtcars$`car name`)  # convert to factor to retain sorted order in plot.

# Diverging Barcharts
ggplot(mtcars, aes(x=`car name`, y=mpg_z, label=mpg_z)) + 
  geom_bar(stat='identity', aes(fill=mpg_type), width=.5)  +
  scale_fill_manual(name="Mileage", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="#00ba38", "below"="#f8766d")) + 
  labs(subtitle="Normalised mileage from 'mtcars'", 
       title= "Diverging Bars") + 
  coord_flip()

```
#### Area Chart
Area charts are typically used to visualize how a particular metric (such as % returns from a stock) performed compared to a certain baseline. Other types of %returns or %change data are also commonly used. The `geom_area()` implements this.

```{r}
data("economics", package = "ggplot2")

# Compute % Returns
economics$returns_perc <- c(0, diff(economics$psavert)/economics$psavert[-length(economics$psavert)])

# Create break points and labels for axis ticks
brks <- economics$date[seq(1, length(economics$date), 12)]
lbls <- lubridate::year(economics$date[seq(1, length(economics$date), 12)])

# Plot
ggplot(economics[1:100, ], aes(date, returns_perc)) + 
  geom_area() + 
  scale_x_date(breaks=brks, labels=lbls) + 
  theme(axis.text.x = element_text(angle=90)) + 
  labs(title="Area Chart", 
       subtitle = "Perc Returns for Personal Savings", 
       y="% Returns for Personal savings", 
       caption="Source: economics")

```

### 3. Ranking
Used to compare the position or performance of multiple items with respect to each other. Actual values matters somewhat less than the ranking.

#### Ordered Bar Chart
Ordered Bar Chart is a Bar Chart that is ordered by the Y axis variable. Just sorting the dataframe by the variable of interest isn’t enough to order the bar chart. In order for the bar chart to retain the order of the rows, the X axis variable (i.e. the categories) has to be converted into a factor.

Let’s plot the mean city mileage for each manufacturer from mpg dataset. First, aggregate the data and sort it before you draw the plot. Finally, the X variable is converted to a factor.

```{r}
cty_mpg <- aggregate(mpg$cty, by=list(mpg$manufacturer), FUN=mean)  # aggregate
colnames(cty_mpg) <- c("make", "mileage")  # change column names
cty_mpg <- cty_mpg[order(cty_mpg$mileage), ]  # sort
cty_mpg$make <- factor(cty_mpg$make, levels = cty_mpg$make)  # to retain the order in plot.
head(cty_mpg, 4)
```
The X variable is now a factor, let’s plot.
```{r}
# Draw plot
ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_bar(stat="identity", width=.5, fill="tomato3") + 
  labs(title="Ordered Bar Chart", 
       subtitle="Make Vs Avg. Mileage", 
       caption="source: mpg") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6))

```
#### Dot Plot

Dot plots emphasizes more on the rank ordering of items with respect to actual values and how far apart are the entities with respect to each other.

```{r}
# Plot
ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_point(col="tomato2", size=3) +   # Draw points
  geom_segment(aes(x=make, 
                   xend=make, 
                   y=min(mileage), 
                   yend=max(mileage)), 
               linetype="dashed", 
               size=0.1) +   # Draw dashed lines
  labs(title="Dot Plot", 
       subtitle="Make Vs Avg. Mileage", 
       caption="source: mpg") +  
  coord_flip()

```

### 4. Distribution
When you have lots and lots of data points and want to study where and how the data points are distributed.

#### Histogram

##### Histogram on a continuous variable
Histogram on a continuous variable can be accomplished using either `geom_bar()` or `geom_histogram()`. When using `geom_histogram()`, you can control the number of bars using the `bins` option. Else, you can set the range covered by each bin using `binwidth`. The value of binwidth is on the same scale as the continuous variable on which histogram is built. Since, `geom_histogram` gives facility to control both number of `bins` as well as `binwidth`, it is the preferred option to create histogram on continuous variables.

```{r}
# Histogram on a Continuous (Numeric) Variable
g <- ggplot(mpg, aes(displ)) + scale_fill_brewer(palette = "Spectral")

g + geom_histogram(aes(fill=class), 
                   binwidth = .1, 
                   col="black", 
                   size=.1) +  # change binwidth
  labs(title="Histogram with Auto Binning", 
       subtitle="Engine Displacement across Vehicle Classes")  

g + geom_histogram(aes(fill=class), 
                   bins=5, 
                   col="black", 
                   size=.1) +   # change number of bins
  labs(title="Histogram with Fixed Bins", 
       subtitle="Engine Displacement across Vehicle Classes") 

```

#### Histogram on a categorical variable
Histogram on a categorical variable would result in a frequency chart showing bars for each category. By adjusting width, you can adjust the thickness of the bars.
```{r}
# Histogram on a Categorical variable
g <- ggplot(mpg, aes(manufacturer))
g + geom_bar(aes(fill=class), width = 0.5) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(title="Histogram on Categorical Variable", 
       subtitle="Manufacturer across Vehicle Classes") 
```
#### Density Plot
```{r}
# Plot
g <- ggplot(mpg, aes(cty))
g + geom_density(aes(fill=factor(cyl)), alpha=0.8) + 
    labs(title="Density plot", 
         subtitle="City Mileage Grouped by Number of cylinders",
         caption="Source: mpg",
         x="City Mileage",
         fill="# Cylinders")

```
#### Box Plot

Box plot is an excellent tool to study the distribution. It can also show the distributions within multiple groups, along with the median, range and outliers if any.

The dark line inside the box represents the median. The top of box is 75%ile and bottom of box is 25%ile. The end points of the lines (aka whiskers) is at a distance of 1.5*IQR, where IQR or Inter Quartile Range is the distance between 25th and 75th percentiles. The points outside the whiskers are marked as dots and are normally considered as extreme points.

Setting `varwidth=T` adjusts the width of the boxes to be proportional to the number of observation it contains.

```{r}
# Plot
g <- ggplot(mpg, aes(class, cty))
g + geom_boxplot(varwidth=T, fill="plum") + 
    labs(title="Box plot", 
         subtitle="City Mileage grouped by Class of vehicle",
         caption="Source: mpg",
         x="Class of Vehicle",
         y="City Mileage")
```
#### Dot + Box Plot
On top of the information provided by a box plot, the dot plot can provide more clear information in the form of summary statistics by each group. The dots are staggered such that each dot represents one observation. So, in below chart, the number of dots for a given manufacturer will match the number of rows of that manufacturer in source data.

```{r}
# plot
g <- ggplot(mpg, aes(manufacturer, cty))
g + geom_boxplot() + 
  geom_dotplot(binaxis='y', 
               stackdir='center', 
               dotsize = .5, 
               fill="red") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(title="Box plot + Dot plot", 
       subtitle="City Mileage vs Class: Each dot represents 1 row in source data",
       caption="Source: mpg",
       x="Class of Vehicle",
       y="City Mileage")

```


