---
title: "understanding data visualization"
author: "Ashish Dutt"
date: "February 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to data visualization
With ever increasing volume of data, it is impossible to tell stories without visualizations. Data visualization is an art of how to turn numbers into useful knowledge. Before the technical implementations of the visualization, let’s see first how to select the right chart type.

An effective chart is one that:

1. Conveys the right information without distorting facts.
2 Is simple but elegant. It should not force you to think much in order to get it.
3. Aesthetics supports information rather that overshadow it.
4. Is not overloaded with information.

### Selecting the Right Chart Type
Primarily, there are 8 types of objectives you may construct plots. So, before you actually make the plot, try and figure what findings and relationships you would like to convey or examine through the visualization. Chances are it will fall under one (or sometimes more) of these 8 categories.

```{r}
knitr::include_graphics("C:\\Users\\ashish.dutt\\Downloads\\choosing-a-good-chart.pdf")

```

In your day-to-day activities, you’ll come across the below listed 7 charts most of the time.

1. Correlation
  + Scatterplot
  + Scatterplot With Encircling
  + Jitter Plot
  + Bubble Plot
  + Animated Bubble Plot
  + Marginal Histogram / Boxplot
2. Deviation
  + Diverging Bars
  + Area Chart
3. Ranking
  + Ordered Bar Chart
  + Dot Plot
  + Slope Chart
4. Distribution
  + Histogram
  + Density Plot
  + Box Plot
  + Dot + Box Plot
  + Tufte Boxplot
  + Violin Plot
5. Composition
  + Waffle Chart
  + Bar Chart
6. Change
  + Time Series Plots
  + Stacked Area Chart
  + Seasonal Plot
7. Groups
  + Dendrogram
  + Clusters
8. Spatial
+ Open Street Map
+ Google Road Map


1. Scatter Plot

**When to use**: Scatter Plot is used to see the relationship between two continuous variables.

In R, a simple scatterplot can be constructed like,

```{r}
library(ggplot2)
ggplot(data=mtcars, aes(x=disp, y=mpg))+
  geom_point()
```
A categorical variable can be used in a scatterplot to show the characterstics of the plotted variables. So, the above code can be modified as,

```{r}
ggplot(data=mtcars, aes(x=disp, y=mpg))+
  geom_point(aes(color=gear))
```

2. Histogram

**When to use**: 
